Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2018-01-15 paper/paper.tex
Add comments on text (except EOB section)

%% macros for comments
\usepackage{color}
%% colors
\definecolor{cyan}{rgb}{0,0.9,0.9}
\definecolor{orange}{rgb}{0.9,0.5,0}
\definecolor{magenta}{rgb}{1,0,1}
\definecolor{purple}{rgb}{0.8,0.4,0.8}
\definecolor{gray}{rgb}{0.8242,0.8242,0.8242}
%%
\newcommand{\bs}[1]{{\textcolor{green}{\texttt{SB: #1}} }}
\newcommand{\bl}[1]{{\textcolor{blue}{\texttt{BL: #1}} }}
\newcommand{\cg}[1]{{\textcolor{orange}{{CG: #1}} }}
\newcommand{\cvdb}[1]{{\textcolor{purple}{\texttt{CVDB: #1}} }}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
%% +++++++++++++++++++++++++++++++++++++++++++++++++++++++
\begin{document}

@@

These results are obtained using Bayesian analyses that compare the data to parameterized waveform models. To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects. This includes descriptions of the point-particle interactions, spins, and tides. Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models. These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order. Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Alternative models exist with more accurate representations of the point particle terms. However, they do not always simultaneously include the spin and tidal effects as well. Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}. In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].

@@

The final step is to interpolate between waveform parameters $\bx$. Previous work on surrogate models focused on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible. Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

@@

This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017}. The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

@@

Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$. The amplitude and phase are smoother, mostly monotonic functions of frequency. This can be seen in Fig.~\ref{fig:hoff} where we show the waveforms for the 32 corners of the 5-dimensional parameter space. Unfortunately, the amplitude and phase still span a wide range of values. The phase, for example, spans about $10^4$~rad between waveforms with different parameters $\bx$ (see Fig.~\ref{fig:hoff}). To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad. This is a difficult requirement to achieve for 5-dimensional interpolation.

@@

For aligned-spin waveforms, we can solve this problem by using the fact that the waveform can be approximated with the analytic \texttt{TaylorF2} waveform $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$. This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
\begin{align}
\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
\label{eq:hdecomp}
\end{align}
These residuals are also shown in Fig.~\ref{fig:hoff}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}). In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior. Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies. In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$. These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work. The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.

@@

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
\end{equation}
where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}. These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.

@@

Finally, we truncate the residuals outside the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}] = [0.00021, 0.07]$. We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$. Although second generation detectors are only marginally sensitive to BNS signals above this frequency, we use the high-frequency cutoff of $Mf_{{\rm trunc}, f}=0.07$ because the amplitude of some of the EOB waveforms is still large enough at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.

To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $S_1$. The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed. The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise.

@@

For the problem here, we have found that global basis functions do not work well. The residuals are small and smooth at low frequencies and large and noisy at high frequencies. With global basis functions, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies. Instead, we find that spline interpolation works significantly better. We use 40 frequency nodes $MF_j$ log-spaced in the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}]$ (see Fig.~\ref{fig:hoff}). We evaluate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at these nodes as discussed below. We then interpolate between these frequencies using third-order splines. The local, third-order polynomials, that are only connected by the requirement of smoothness, do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.

@@

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At $\sim10$~minutes--1~day per EOB waveform on a standard CPU, this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is Gaussian Process Regression (GPR)~\cite{RasmussenWilliams2006}.

@@

@@

An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD and no points are repeated in any dimension. This avoids wasting samples when one of the parameters has much less of an influence than the other parameters. As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies. However, during the merger the tidal parameters can be more important than the other parameters. One therefore might want more samples for the tidal parameters and less samples for the other parameters. However, due to the expense of the waveforms we must use the same set of waveforms for low and high frequencies. An LHD with its noncollapsing property bypasses this problem.

@@

With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion and update the surrogate with each new waveform. This method is sometimes referred to as uncertainty sampling~\cite{BrochuCoraDeFreitas2010}. The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$:
\begin{equation}
\epsilon_{\rm RMS}(\bx) = \sqrt{\frac{1}{N} \sum_{ \substack{j \\ MF_j \le 0.03} } [\sigma_{\Delta\Phi_j}(\bx)]^2},
\label{eq:errorrms}
\end{equation}
where $\sigma_{\Delta\Phi_j}(\bx)$ is the GPR estimate (Eq.~\eqref{eq:var}) of the uncertainty in $\Delta\Phi(MF_j, \bx)$. There are many possible ways to construct a scalar that estimates overall waveform error. An alternative would be an approximation to the mismatch discussed in Sec.~\ref{sec:accuracy}. However, the quantity $\epsilon_{\rm RMS}(\bx)$ describes the most important quantity, phase, and is fast to compute. For an initial training set with $N$ samples, $\epsilon_{\rm RMS}(\bx)$ has $\sim N$ local maxima located in the voids between samples $\bx$. To find the global maximum, we use a basin hopping algorithm~\cite{WalesDoye1998, scipy:basinhopping} to avoid getting stuck in local maxima.

@@

@@

>>>>

%% macros for comments
\usepackage{color}
%% colors
\definecolor{cyan}{rgb}{0,0.9,0.9}
\definecolor{orange}{rgb}{0.9,0.5,0}
\definecolor{magenta}{rgb}{1,0,1}
\definecolor{purple}{rgb}{0.8,0.4,0.8}
\definecolor{gray}{rgb}{0.8242,0.8242,0.8242}
%%
\newcommand{\MP}[1]{{\textcolor{blue}{\texttt{MP: #1}} }}
\newcommand{\bs}[1]{{\textcolor{green}{\texttt{SB: #1}} }}
\newcommand{\bl}[1]{{\textcolor{blue}{\texttt{BL: #1}} }}
\newcommand{\cg}[1]{{\textcolor{orange}{{CG: #1}} }}
\newcommand{\cvdb}[1]{{\textcolor{purple}{\texttt{CVDB: #1}} }}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
%% +++++++++++++++++++++++++++++++++++++++++++++++++++++++
\begin{document}

@@

These results are obtained using Bayesian analyses that compare the data to parameterized waveform models. To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects. This includes descriptions of the point-particle interactions, spins, and tides. Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models. These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order. Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Alternative models exist with more accurate representations of the point particle terms. \MP{What are these alternative models besides EOB?} However, they do not always simultaneously include the spin and tidal effects as well. Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}. In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].

@@

The final step is to interpolate between waveform parameters $\bx$. Previous work on surrogate models focused on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible \MP{I think that this is too strong; Blackman17a also perform 5D TP interpolation which contradicts this.}. Recent work on optimally choosing waveforms for both EOB\MP{We used IMRPhenomD}~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

@@

This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. \MP{Is it really true that this is a significant speedup? FFT may still be cheaper than surrogate waveform generation.} It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} \MP{TODO: add citation to PhenomP ROQ paper here and later} and multi-banding~\cite{VinciguerraVeitchMandel2017}. The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

@@

Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$. The amplitude and phase are smoother, mostly monotonic functions of frequency \MP{the phase is not guaranteed to be monotonic}. This can be seen in Fig.~\ref{fig:hoff} where we show the waveforms for the 32 corners of the 5-dimensional parameter space. Unfortunately, the amplitude and phase still span a wide range of values. The phase, for example, spans about $10^4$~rad between waveforms with different parameters $\bx$ (see Fig.~\ref{fig:hoff}). To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad. This is a difficult requirement to achieve for 5-dimensional interpolation.

@@

For aligned-spin waveforms, we can solve this problem by using the fact that the waveform can be approximated with the analytic \texttt{TaylorF2} waveform $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$. This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
\begin{align}
\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
\label{eq:hdecomp}
\end{align}
These residuals are also shown in Fig.~\ref{fig:hoff}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}). In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior. Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies. In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$. \MP{TODO: replace $S_i$ by $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$.} These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work. The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.

@@

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
\end{equation}
where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}. These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}. \MP{add git hash? LAL's TF2 will change in the future when new terms become available.}

@@

Finally, we truncate the residuals outside the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}] = [0.00021, 0.07]$. We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$. Although second generation detectors are only marginally sensitive to BNS signals above this frequency, we use the high-frequency cutoff of $Mf_{{\rm trunc}, f}=0.07$ because the amplitude of some of the EOB waveforms is still large enough at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain. \MP{Say that Mf=0.07 is a (generous?) upper bound on the NR BNS merger frequency and thus a safe choice.}

To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $S_1$. The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed. The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise. 
\MP{The discussion of Fig.~\ref{fig:dhofs} appears twice in the text; here and one section earlier.}

@@

For the problem here, we have found that global basis functions do not work well. \MP{EIM did not work well; is this also true for the SVD basis and projection coefficient method?} The residuals are small and smooth at low frequencies and large and noisy at high frequencies. With global basis functions, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies. Instead, we find that spline interpolation works significantly better. We use 40 frequency nodes $MF_j$ log-spaced in the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}]$ (see Fig.~\ref{fig:hoff}). We evaluate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at these nodes as discussed below. We then interpolate between these frequencies using third-order splines. The local, third-order polynomials, that are only connected by the requirement of smoothness, do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.

@@

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At $\sim10$~minutes--1~day per EOB waveform on a standard CPU \MP{For which $f_\min$ are these times given?}, this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is Gaussian Process Regression (GPR)~\cite{RasmussenWilliams2006}.

@@

\MP{Say that we use sparse frequency grids based on power-law $\Delta f$ in the evaluation of the surrogate; perhaps describe in more detail how the model is evaluated in the LAL code.}
\MP{Discuss linear vs cubic splines if we decide to use them. Tradeoff in accuracy and speed.}

@@

An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD and no points are repeated in any dimension. This avoids wasting samples when one of the parameters has much less of an influence than the other parameters. As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies. However, during the merger the tidal parameters can be more important than the other parameters. One therefore might want more samples for the tidal parameters and less samples for the other parameters. However, due to the expense of the waveforms we must use the same set of waveforms for low and high frequencies. An LHD with its noncollapsing property bypasses this problem. 
\MP{Note (perhaps in a footnotes) that it is possible to have separate surrogates for low and high frequency which are then joined smoothly as done for \texttt{SEOBNRv4\_ROM}.}

@@

With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion and update the surrogate with each new waveform. This method is sometimes referred to as uncertainty sampling~\cite{BrochuCoraDeFreitas2010}. The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$: \MP{Add reason why higher frequencies are not included in this error measure.}
\begin{equation}
\epsilon_{\rm RMS}(\bx) = \sqrt{\frac{1}{N} \sum_{ \substack{j \\ MF_j \le 0.03} } [\sigma_{\Delta\Phi_j}(\bx)]^2},
\label{eq:errorrms}
\end{equation}
where $\sigma_{\Delta\Phi_j}(\bx)$ is the GPR estimate (Eq.~\eqref{eq:var}) of the uncertainty in $\Delta\Phi(MF_j, \bx)$. There are many possible ways to construct a scalar that estimates overall waveform error. An alternative would be an approximation to the mismatch discussed in Sec.~\ref{sec:accuracy}. However, the quantity $\epsilon_{\rm RMS}(\bx)$ describes the most important quantity, phase, and is fast to compute. For an initial training set with $N$ samples, $\epsilon_{\rm RMS}(\bx)$ has $\sim N$ local maxima located in the voids between samples $\bx$. To find the global maximum, we use a basin hopping algorithm~\cite{WalesDoye1998, scipy:basinhopping} to avoid getting stuck in local maxima.

@@

\MP{Add cost of complex exponential sincos function: look at profiling results}
\MP{Potentially, explore approximation to sincos for further speedup; need to guarantee that errors don't add up; range reduction is important for accuracy}

@@

\subsection{Parameter estimation}

\MP{TODO: add 2D density plot of which quantities? masses, spins, lambdas?
or mass-ratio, eff spin, eff lambda?}
